Joint maximum-likelihood of phylogenies and ancestral states is not consistent

### 2018-08-06 appendix.tex
:male-factory-worker:

Consider the InvFels tree $\tau_1$ with arbitrary fidelities, i.e., $t=\{x_1,y_1,x_2,y_2,w\}$.
Next we show that the likelihood $\ell_{\tau_1,t^*}(\tau_1, t; \boldsymbol\xi)$ remains unchanged if $x_1$ and $x_2$ are exchanged or if $y_1$ and $y_2$ are.
Using the Hadamard transform, we calculate the generating probabilities on the InvFels tree.
For site split $\emptyset$,
\begin{align*}
    \Pr(\siteSplitRV=\emptyset\mid \tau_1, t) & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  x_1y_1w + x_1y_2w + y_1x_2w + x_2y_2w + x_1y_1x_2y_2) \\
                                              & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  w[x_1y_1 + x_1y_2 + y_1x_2 + x_2y_2] + x_1y_1x_2y_2) \\
                                              & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  w[x_1 + x_2][y_1 + y_2] + x_1y_1x_2y_2),
\end{align*}
and this probability is unchanged when $x_1$ is exchanged with $x_2$ and $y_1$ is exchanged with $y_2$.
Similarly, for site split $\{1,3\}$,
\[
    \Pr(\siteSplitRV=\{1,3\}\mid \tau_1, t) = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 -  w[x_1 + x_2][y_1 + y_2] + x_1y_1x_2y_2),
\]
which also is invariant to exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$.

@@

Exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$ does not change the value of the log-likelihood $\ell_{\tau_1,t^*}(\tau_1, t; \boldsymbol\xi)$.
Thus, we can reduce the number of candidate likelihoods we need to search by, without loss of generality, assuming $x_2 \ge x_1$ and $y_2 \ge y_1$, with these conditional likelihoods given in Table~\ref{tab:likelihoods} after maximizing over ancestral state splits.

@@

\begin{proof}
We proceed by analyzing the gradients of each component of the likelihood function.
We first focus on the marginal log-likelihood
\[
\shannonDivergence_{\tau^*,t^*}(\tau_1,t) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \log \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, t).
\]
By Gibbs' inequality, this function attains a unique maximum when
\[
\Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, t) = \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)
\]
for all $j=1,\ldots,\nSiteSplits$ with $t^*=\{x^*,y^*,x^*,y^*,y^*\}$.
We show that this implies $t=t^*$.
Note that exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$ does not affect the marginal log-likelihood, and, as it has a unique maximum, we have $x_1=x_2=x$ and $y_1=y_2=y$ at this maximum.
Referring to the top of Table~\ref{tab:gen-sitepatprob}, we set up simultaneous equations as follows.
The sum of generating probabilities for site splits $\{3\}$ and $\{1,2\}$ yields
\[
2-2(x^*)^2 = 2-2x^2
\]
and for $\{2\}$ and $\{1,2\}$ yields
\[
2-2(y^*)^2 = 2-2y^2.
\]
implying $x=x^*$ and $y=y^*$.
Setting these parameters to their optima, the difference of generating probabilities for site splits $\emptyset$ and $\{1,3\}$ yields
\[
8x^*(y^*)^2 = 8x^*y^*w
\]
implying $w=y^*$.
Thus the unique maximum is $\hat{t}=\{x^*,y^*,x^*,y^*,y^*\}$, and consequently
\[
\nabla_t \shannonDivergence_{\tau^*,t^*}(\tau_1,t) \mid_{t=t^*} = 0.
\]
We now show that the gradient of the marginal log-likelihood is a decreasing function on $(0,1]$ in each variable.
The marginal log-likelihood as a function of $x_1$ is
\begin{align*}
\shannonDivergence(x_1) = \sum_j p_{\siteSplit_j} \log(c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1) - \log 8
\end{align*}
where $c_{x_1,\siteSplit_j}$ and $a_{x_1,\siteSplit_j}$ are functions of $y_1,x_2,y_2,$ and $w$.
For example, reading from Table~\ref{tab:gen-sitepatprob},
\[
c_{x_1,\emptyset} = 1+y_1y_2+wx_2y_1+wx_2y_2
\]
and
\[
a_{x_1,\emptyset} = x_2+wy_1+wy_2+x_2y_1y_2
\]
with similar definitions for the remaining site splits.
By the assumption of $0 < x^*, y^* < 1$, we have $p_{\siteSplit_j} > 0$ for $j=1,\ldots,\nSiteSplits$, therefore we assume $c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1 \neq 0$ for all $j$ otherwise we obtain a zero value for our likelihood, i.e., $\shannonDivergence\rightarrow -\infty$.
In other words, we make the assumption that $x_1,y_1,x_2 < 1$, precluding all cases where $c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1 = 0$ for some $j$.
This assumption is reasonable as the likelihood is never maximized at these values where $c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1 = 0$.
Moreover, we have
\[
\Pr(\siteSplitRV=\siteSplit_j \mid \tau_1,\{x_1,y_1,x_2,y_2,w\}) = \frac{1}{8} (c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1),
\]
which, being a probability, is greater than or equal to zero.
Consider the derivative with respect to $x_1$
\begin{align*}
\frac{d}{dx_1} \shannonDivergence(x_1) = \sum_j \frac{p_{\siteSplit_j}a_{x_1,\siteSplit_j}}{c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1}
\end{align*}
and focus on the $j$th term in the summation
\[
f_j(x_1) := \frac{p_{\siteSplit_j}a_{x_1,\siteSplit_j}}{c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1}.
\]
%das: gah, had to roll back this line and add some stuff above since there are cases where some probabilities are 0. Maybe I'm being too careful? I don't think it's an issue, just an edge case that could induce wonkiness, so I included an assumption.
Regardless of $a_{x_1,\siteSplit_j} \lessgtr 0$, $f_j$ is a decreasing function of $x_1$ as $c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1 > 0$.
The remaining variables follow similar arguments.

@@

We now focus on the conditional log-likelihood.
Assume $\emptyset$ is the most likely ancestral state split for all site splits.
In this case, the conditional log-likelihood has no ambiguity, and has terms given as the log of those in Table~\ref{tab:likelihoods-restricted}.
We focus on the derivative with respect to $x_1$, aggregating all terms involving $x_2,y_1,y_2,$ and $w$ into a constant term.
The conditional log-likelihood
\[
\ell^C_{\tau^*,t^*}(\tau_1, t; \boldsymbol\xi_0) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log \Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV = \siteSplit_j, \tau_1, t),
\]
using Tables~\ref{tab:gen-sitepatprob} and~\ref{tab:likelihoods-restricted}, is
\begin{align*}
\ell^C(x_1; \boldsymbol\xi_0) &= \left(\frac{1}{2}+\frac{1}{2}x^*(y^*)^2\right)\log(1+x_1) + \left(\frac{1}{2}-\frac{1}{2}x^*(y^*)^2\right)\log(1-x_1)\\
&\qquad + \left(\frac{1}{2}+\frac{1}{2}(y^*)^2\right)\log(1+y_1) + \left(\frac{1}{2}-\frac{1}{2}(y^*)^2\right)\log(1-y_1)\\
&\qquad + \left(\frac{1}{2}+\frac{1}{2}x^*(y^*)^2\right)\log(1+x_2) + \left(\frac{1}{2}-\frac{1}{2}x^*(y^*)^2\right)\log(1-x_2)\\
&\qquad + \log(1+y_2) + \log(1+w) - \log 32\\
&:= \left(\frac{1}{2}+\frac{1}{2}x^*(y^*)^2\right)\log(1+x_1) + \left(\frac{1}{2}-\frac{1}{2}x^*(y^*)^2\right)\log(1-x_1)\\
&\qquad + C_{x_1}(x_2,y_1,y_2,w).
\end{align*}
where $C_{x_1}$ is a function of $y_1,x_2,y_2,$ and $w$.
The derivative with respect to $x_1$ is
\[
\frac{d}{dx_1} \ell^C(x_1; \boldsymbol\xi_0) = \frac{-x_1+x^*(y^*)^2}{1-x_1^2},
\]
which is negative for all $x_1 \ge x^*$.
This same process applied to $x_2$ and $y_1$ shows the gradient of the full likelihood is negative for $x_2 \ge x^*$ and $y_1 \ge y^*$.

The variables $y_2$ and $w$ require the following.
Consider the conditional log-likelihood with respect to $w$
\[
\ell^C(w; \boldsymbol\xi_0) := \log(1+w) + C_{w}(x_1,y_1,x_2,y_2)
\]
with derivative with respect to $w$
\[
\frac{d}{dw} \ell^C(w; \boldsymbol\xi_0) = \frac{1}{1+w},
\]
which is positive for all $w \ge 0$.
As the gradient of the marginal log-likelihood is a decreasing function in $w$ and has a zero crossing at $y^*$, the addition of the derivative of the conditional log-likelihood makes the gradient of the full log-likelihood positive for all $w \le y^*$.
The same argument holds for $y_2$, yielding the constraints
\[
\hat{x}_1,\hat{x}_2 \in (0, x^*], \ \hat{y}_1 \in (0, y^*], \text{and}\ \hat{y}_2,\hat{w} \in [y^*, 1].
\]
\end{proof}

Since $x^*, y^* < 1$, the assumption that $x_1,y_1,x_2 < 1$ does not contradict with the statement of Lemma~\ref{lemma:bounds}.
We now prove the main result.

>>>>

Consider the InvFels tree $\tau_1$ with arbitrary fidelities, i.e., $t=\{x_1,y_1,x_2,y_2,w\}$.
Next we show that the likelihood $\ell_{\tau_1,t^*}(\tau_1, t; \boldsymbol\xi)$ remains unchanged if $x_1$ and $x_2$ are exchanged or if $y_1$ and $y_2$ are.
Although this property should not be surprising due to symmetry, we write it out for completeness.
Using the Hadamard transform, we calculate the generating probabilities on the InvFels tree.
For site split $\emptyset$,
\begin{align*}
    \Pr(\siteSplitRV=\emptyset\mid \tau_1, t) & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  x_1y_1w + x_1y_2w + y_1x_2w + x_2y_2w + x_1y_1x_2y_2) \\
                                              & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  w[x_1y_1 + x_1y_2 + y_1x_2 + x_2y_2] + x_1y_1x_2y_2) \\
                                              & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  w[x_1 + x_2][y_1 + y_2] + x_1y_1x_2y_2),
\end{align*}
and this probability is unchanged when $x_1$ is exchanged with $x_2$ and $y_1$ is exchanged with $y_2$.
Similarly, for site split $\{1,3\}$,
\[
    \Pr(\siteSplitRV=\{1,3\}\mid \tau_1, t) = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 -  w[x_1 + x_2][y_1 + y_2] + x_1y_1x_2y_2),
\]
which also is invariant to exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$.

@@

Thus exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$ does not change the value of the log-likelihood $\ell_{\tau_1,t^*}(\tau_1, t; \boldsymbol\xi)$.
Therefore we can reduce the number of candidate likelihoods we need to search by, without loss of generality, assuming $x_2 \ge x_1$ and $y_2 \ge y_1$, with these conditional likelihoods given in Table~\ref{tab:likelihoods} after maximizing over ancestral state splits.

@@

\begin{proof}
We proceed by analyzing the gradients of each component of the likelihood function.
We first focus on the marginal log-likelihood
\[
\shannonDivergence_{\tau^*,t^*}(\tau_1,t) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \log \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, t).
\]
By Gibbs' inequality, this function attains a unique maximum when
\[
\Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, t) = \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)
\]
for all $j=1,\ldots,\nSiteSplits$ with $t^*=\{x^*,y^*,x^*,y^*,y^*\}$ and $\tau^*$ being the InvFels tree $\tau_1$.
We show that this implies $t=t^*$.
Note that exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$ does not affect the marginal log-likelihood, and, as it has a unique maximum, we have $x_1=x_2=x$ and $y_1=y_2=y$ at this maximum.
Referring to the top of Table~\ref{tab:gen-sitepatprob}, we set up simultaneous equations as follows.
The sum of generating probabilities for site splits $\{3\}$ and $\{1,2\}$ yields
\[
2-2(x^*)^2 = 2-2x^2
\]
and for $\{2\}$ and $\{1,2\}$ yields
\[
2-2(y^*)^2 = 2-2y^2.
\]
implying $x=x^*$ and $y=y^*$.
Setting these parameters to their optima, the difference of generating probabilities for site splits $\emptyset$ and $\{1,3\}$ yields
\[
8x^*(y^*)^2 = 8x^*y^*w
\]
implying $w=y^*$.
Thus the unique maximum is $\hat{t}=\{x^*,y^*,x^*,y^*,y^*\}$, and consequently
\[
\nabla_t \shannonDivergence_{\tau^*,t^*}(\tau_1,t) \mid_{t=t^*} = 0.
\]
We now show that the gradient of the marginal log-likelihood is a decreasing function on $(0,1]$ in each variable.
The marginal log-likelihood as a function of $x_1$ is
\begin{align*}
\shannonDivergence(x_1) = \sum_j p_{\siteSplit_j} \log(c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1) - \log 8
\end{align*}
where $c_{x_1,\siteSplit_j}$ and $a_{x_1,\siteSplit_j}$ are functions of $y_1,x_2,y_2,$ and $w$.
For example, reading from the middle panel of Table~\ref{tab:gen-sitepatprob},
\[
c_{x_1,\emptyset} = 1+y_1y_2+wx_2y_1+wx_2y_2
\]
and
\[
a_{x_1,\emptyset} = x_2+wy_1+wy_2+x_2y_1y_2
\]
with similar definitions for the remaining site splits.

%EM Could you restructure this paragraph with a topic-sentence statement of the assumption we want to make, then evidence that it's ok? It roves around a good bit.
By the assumption of $0 < x^*, y^* < 1$, we have $p_{\siteSplit_j} > 0$ for $j=1,\ldots,\nSiteSplits$, therefore we assume $c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1 \neq 0$ for all $j$ otherwise we obtain a zero value for our likelihood, i.e., $\shannonDivergence\rightarrow -\infty$.
In other words, we make the assumption that $x_1,y_1,x_2 < 1$, precluding all cases where $c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1 = 0$ for some $j$.
This assumption is reasonable as the likelihood is never maximized at these values where $c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1 = 0$.

Moreover, we have
\[
\Pr(\siteSplitRV=\siteSplit_j \mid \tau_1,\{x_1,y_1,x_2,y_2,w\}) = \frac{1}{8} (c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1),
\]
which, being a probability, is greater than or equal to zero.
Consider the derivative with respect to $x_1$
\begin{align*}
\frac{d}{dx_1} \shannonDivergence(x_1) = \sum_j \frac{p_{\siteSplit_j}a_{x_1,\siteSplit_j}}{c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1}
\end{align*}
and focus on the $j$th term in the summation
\[
f_j(x_1) := \frac{p_{\siteSplit_j}a_{x_1,\siteSplit_j}}{c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1}.
\]
%das: gah, had to roll back this line and add some stuff above since there are cases where some probabilities are 0. Maybe I'm being too careful? I don't think it's an issue, just an edge case that could induce wonkiness, so I included an assumption.
Regardless of $a_{x_1,\siteSplit_j} \lessgtr 0$, $f_j$ is a decreasing function of $x_1$ as $c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1 > 0$.
The remaining variables follow similar arguments.

@@

We now focus on the conditional log-likelihood.
%EM These "Assume"s feel frightening to the reader. Can we replace with "Under our stated hypothesis"?
Assume $\emptyset$ is the most likely ancestral state split for all site splits.
In this case, the conditional log-likelihood has no ambiguity, and has terms given as the log of those in Table~\ref{tab:likelihoods-restricted}.
We focus on the derivative with respect to $x_1$, aggregating all terms involving $x_2,y_1,y_2,$ and $w$ into a constant term.
The conditional log-likelihood
\[
\ell^C_{\tau^*,t^*}(\tau_1, t; \boldsymbol\xi_0) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log \Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV = \siteSplit_j, \tau_1, t),
\]
using Tables~\ref{tab:gen-sitepatprob} and~\ref{tab:likelihoods-restricted}, is
\begin{align*}
\ell^C(x_1; \boldsymbol\xi_0) &= \left(\frac{1}{2}+\frac{1}{2}x^*(y^*)^2\right)\log(1+x_1) + \left(\frac{1}{2}-\frac{1}{2}x^*(y^*)^2\right)\log(1-x_1)\\
&\qquad + \left(\frac{1}{2}+\frac{1}{2}(y^*)^2\right)\log(1+y_1) + \left(\frac{1}{2}-\frac{1}{2}(y^*)^2\right)\log(1-y_1)\\
&\qquad + \left(\frac{1}{2}+\frac{1}{2}x^*(y^*)^2\right)\log(1+x_2) + \left(\frac{1}{2}-\frac{1}{2}x^*(y^*)^2\right)\log(1-x_2)\\
&\qquad + \log(1+y_2) + \log(1+w) - \log 32\\
&:= \left(\frac{1}{2}+\frac{1}{2}x^*(y^*)^2\right)\log(1+x_1) + \left(\frac{1}{2}-\frac{1}{2}x^*(y^*)^2\right)\log(1-x_1)\\
&\qquad + C_{x_1}(x_2,y_1,y_2,w).
\end{align*}
where $C_{x_1}$ is a function of $y_1,x_2,y_2,$ and $w$.
The derivative with respect to $x_1$ is
\[
\frac{d}{dx_1} \ell^C(x_1; \boldsymbol\xi_0) = \frac{-x_1+x^*(y^*)^2}{1-x_1^2},
\]
which is negative for all $x_1 \ge x^*$.
This same process applied to $x_2$ and $y_1$ shows the gradient of the full likelihood is negative for $x_2 \ge x^*$ and $y_1 \ge y^*$.

The variables $y_2$ and $w$ require the following.
Consider the conditional log-likelihood with respect to $w$
\[
\ell^C(w; \boldsymbol\xi_0) := \log(1+w) + C_{w}(x_1,y_1,x_2,y_2)
\]
with derivative with respect to $w$
\[
\frac{d}{dw} \ell^C(w; \boldsymbol\xi_0) = \frac{1}{1+w},
\]
which is positive for all $w \ge 0$.
%EM Does the zero crossing need additional argument? I'm guessing that you get this from the gradient-is-zero thing from the unique-maximum argument above. But that's just that it's zero when we are doing maximization of all of the variables at once. Here aren't we wanting to make an argument for lots of x_i and y_i?
As the gradient of the marginal log-likelihood is a decreasing function in $w$ and has a zero crossing at $y^*$, the addition of the derivative of the conditional log-likelihood makes the gradient of the full log-likelihood positive for all $w \le y^*$.
The same argument holds for $y_2$, yielding the constraints
\[
\hat{x}_1,\hat{x}_2 \in (0, x^*], \ \hat{y}_1 \in (0, y^*], \text{and}\ \hat{y}_2,\hat{w} \in [y^*, 1].
\]
\end{proof}

%EM Are we saying that Lemma 3 applies in this case? I think so but it could be more specific.
Since $x^*, y^* < 1$, the assumption that $x_1,y_1,x_2 < 1$ does not contradict with the statement of Lemma~\ref{lemma:bounds}.
We now prove the main result.
